\bf Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-15 manuscript/main.tex
proposition 1 and 2 relabling

	\begin{proposition}\label{proposition:generalized_rayleigh_quotient}
At the $l$-th iteration of the PLS algorithm, let $
( \hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top )^\top
$ denote the coefficients 
of $	\hat{\xi}$, 
defined in \eqref{def:maximizer_squared_empirical_cov},
which is
the projected unit-norm direction that maximizes the squared empirical covariance.

 These coefficients can be obtained by solving the following generalized eigenvalue problem
\begin{align}
	\left(
	\hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top
	\right)^\top
	=
	\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
	\quad \text{subject to} \quad \boldsymbol{\xi}^\top B^\ast \boldsymbol{\xi} = 1,
	\label{eq: Regularized generalized rayleigh quotient equation}
\end{align}
where
$$
V := \frac{1}{n^2} (B^\ast \Theta^\top \mathbf{y})(B^\ast \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)},
$$

\end{proposition}
Since $V$ is a rank 1 matrix, the solution of the generalized eigenvalue problem is easily obtained by matrix multiplication and normalization:
	\begin{proposition} \label{proposition:generalized_rayleigh_quotient_linear}
	We can compute  $\hat{\mathbf{d}}_1, \ldots, \hat{\mathbf{d}}_K$ and $\hat{\boldsymbol{\zeta}}$ in Proposition \ref{proposition:generalized_rayleigh_quotient} as 
\begin{equation*}
	\hat{\mathbf{d}}_j  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	\Theta_j^\top \mathbf{y},
	\quad
	%%%%%%%%%%%%%%%%
	\hat{\boldsymbol{\zeta}}  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	Z^\top \mathbf{y}.
\end{equation*}
\end{proposition}
The expressions above represent matrix–vector multiplications normalized by a common factor. 
Proofs of Propositions \ref{proposition:generalized_rayleigh_quotient} and \ref{proposition:generalized_rayleigh_quotient_linear} are provided in Appendix~\ref{section:proof:proposition:generalized_rayleigh_quotient_linear_system_unregularized}.
Although the unnormalized coefficients are computed separately for the functional and scalar part of the PLS component direction, the normalization step couples them, allowing the components to influence each other. This enables the procedure to account for the correlation between functional and scalar components.

	\subsubsection{Proposed Approach: Regularized and Smoothed Computation} \label{sec: Regularization of PLS Components}
	Although Proposition \ref{proposition:generalized_rayleigh_quotient} provides an efficient and principled way to compute the PLS components $\xi$, their functional part $\psi=(\psi^{(1)}, \ldots, \psi^{(K)})$ may not be smooth, which can lead to overfitting and unstable predictions. Thus, in this section, we provide a modification of Proposition  \ref{proposition:generalized_rayleigh_quotient} that yields regularized (smoothed) PLS components.

@@

	Here, $\lambda_k$ $(k=1, \ldots, K)$ is a positive smoothing parameter that controls the trade-off between the goodness of fit and the amount of smoothness in $\psi^{(k)}$. Smaller $\lambda_k$ produces PLS components that fit more closely to the observed data, but selecting too small value may cause overfitting. The estimated PLS component reverts back to the unregularized version of Proposition \ref{proposition:generalized_rayleigh_quotient} when $\lambda_k=0$ for all $k$. On the other hand, larger $\lambda_k$ places more emphasis on smoothing $\psi^{(k)}$. In the limit $\lambda_k \rightarrow \infty$, each of the functions is forced to be of the form $\psi^{(k)}(t)=a+bt$ for some constants $a$ and $b$. In practice, $\{\lambda_k\}_{k=1}^K$ as well as the number of PLS components $L$ to be estimated can be chosen by cross-validation based on a certain predictive performance criterion (e.g., mean squared error).

@@

	\subsection{Proof of Proposition \ref{proposition:generalized_rayleigh_quotient_linear_system_unregularized}}\label{section:proof:proposition:generalized_rayleigh_quotient_linear_system_unregularized}

@@

	presented in Proposition \ref{proposition:generalized_rayleigh_quotient_linear_system_unregularized} is equivalent to the following constrained optimization problem on the Euclidean space:

@@

	Define the Lagrangian:
	\[
	\mathcal{L}(\{\mathbf{d}_j\}, \boldsymbol{\zeta}, \lambda) 
	:= 
	f(\{\mathbf{d}_j\}, \boldsymbol{\zeta}) - \lambda g(\{\mathbf{d}_j\}, \boldsymbol{\zeta}).
	\]
	The constraint function $g$ is continuously differentiable, and its gradient vanishes only at the origin. Since the origin is not feasible (the constraint requires norm 1), the gradient of $g$ is nonzero at all feasible points. Thus, the linear independence constraint qualification (LICQ) holds, ensuring that the KKT conditions are necessary for local maxima (see, for example, Theorem 5.3.1 in \citealt{bazaraaNonlinearProgrammingTheory2006}).
	Taking gradients of $\mathcal{L}$ and setting them to zero yields:
	\begin{align*}
		\nabla_{\mathbf{d}_j} \mathcal{L} &= 
		\frac{2}{n^2} 
		\left(
		\sum_{\ell=1}^K \mathbf{y}^\top \Theta_\ell B \mathbf{d}_\ell + \mathbf{y}^\top Z \boldsymbol{\zeta}
		\right)
		B \Theta_j^\top \mathbf{y}
		- 2\lambda B \mathbf{d}_j = 0, \quad \forall j, 
		\numberthis \label{lag_grad_d}
		\\
		\nabla_{\boldsymbol{\zeta}} \mathcal{L} &= 
		\frac{2}{n^2} 
		\left(
		\sum_{\ell=1}^K \mathbf{y}^\top \Theta_\ell B \mathbf{d}_\ell + \mathbf{y}^\top Z \boldsymbol{\zeta}
		\right)
		Z^\top \mathbf{y}
		- 2\lambda \boldsymbol{\zeta} = 0.
		\numberthis \label{lag_grad_zeta}
	\end{align*}
	By pre-multiplying equation \eqref{lag_grad_d} by $B^{-1}$ and observing that the scalar quantity $n^{-2}\sum_{\ell=1}^K \mathbf{y}^\top \Theta_\ell B \mathbf{d}_\ell + \mathbf{y}^\top Z \boldsymbol{\zeta}$ appears in both \eqref{lag_grad_d} and \eqref{lag_grad_zeta}, the gradient conditions together imply that any local optimum must be  scalar multiples of
	$
	\Theta_1^\top \mathbf{y}, \ldots, \Theta_K^\top \mathbf{y}, Z^\top \mathbf{y},
	$
	all scaled by the same factor. There exists only one set of vectors  that also satisfies the primal equality constraint, namely the normalized version of the vector set above. The global maximizer is therefore given by
	\begin{equation*}
		\mathbf{d}_j^\ast = \frac{1}{\sqrt{\sum_{j=1}^K 
				\mathbf{y}^\top
				\Theta_j
				B \Theta_j^\top \mathbf{y} 
				+ 
				\mathbf{y}^\top Z Z^\top \mathbf{y}}}
		\Theta_j^\top \mathbf{y},
		\quad
		%%%%%%%%%%%%%%%%
		\boldsymbol{\zeta}^\ast = \frac{1}{\sqrt{\sum_{j=1}^K 
				\mathbf{y}^\top
				\Theta_j
				B \Theta_j^\top \mathbf{y} 
				+ 
				\mathbf{y}^\top Z Z^\top \mathbf{y}}}
		Z^\top \mathbf{y},
	\end{equation*}
	which are the same as  $\hat{\mathbf{d}}_1, \ldots, \hat{\mathbf{d}}_K$ and $\hat{\boldsymbol{\zeta}}$. This completes the proof of  Proposition \ref{proposition:generalized_rayleigh_quotient_linear_system_unregularized}.

	\subsection{Proof of Propositions \ref{proposition:generalized_rayleigh_quotient} and \ref{Proposition: Regularized generalized rayleigh quotient} }\label{section:proof:eigenproblem_derivation}
	%
	%
	\begin{align*}
		\widehat{\textnormal{Cov}} (
		\langle \widetilde{\mathbf{W}}, \xi \rangle_{\mathcal{H}}, Y 
		) 
		&= 
		\frac{1}{n} \sum \limits_{i=1}^n 
		y_i
		\langle \widetilde{\mathbf{W}}_i, \xi \rangle_\mathcal{H} \\
		&= 
		\frac{1}{n} 
		\sum \limits_{i=1}^n
		y_i
		\langle \widetilde{X_i}, \psi \rangle_\mathcal{F}  
		+  
		\frac{1}{n} 
		\sum \limits_{i=1}^n  
		y_i
		\langle \mathbf{Z}_i, 
		\boldsymbol{\theta} 
		\rangle 
		\\
		&=  \frac{1}{n} \sum \limits_{i=1}^n  \left\{ \sum \limits_{k=1}^K \int_0^1 X_i(t_k) \psi^{(k)}(t_k) dt_k \right\} Y_i +  \frac{1}{n} \sum \limits_{i=1}^n \mathbf{Z}_i^\top \boldsymbol{\theta} Y_i \\
		&=  \frac{1}{n} \sum \limits_{i=1}^n \left[ \sum \limits_{k=1}^K \int_0^1 \left\{\sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t) \right\} \left\{\sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t) \right\} dt_k \right] Y_i + \frac{1}{n} \sum \limits_{i=1}^n \mathbf{Z}_i^\top \boldsymbol{\theta} Y_i \\
		&= n^{-1} \widetilde{\mathbf{y}}^\top \left[ \sum \limits_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \left\{ \int_0^1 \mathbf{b}^{(k)}(t_k) \mathbf{b}^{(k)}(t_k)^\top dt_k \right\} \mathbf{a}^{(k)} \right] + n^{-1}\widetilde{\mathbf{y}}^\top \mathbf{Z} \boldsymbol{\theta} \\
		&= n^{-1} \widetilde{\mathbf{y}}^\top (\widetilde{C} J \mathbf{a}) + n^{-1}\widetilde{\mathbf{y}}^\top \mathbf{Z} \boldsymbol{\theta} \\
		&= n^{-1} \widetilde{\mathbf{y}}^\top (\widetilde{C}J\mathbf{a} + \mathbf{Z} \boldsymbol{\theta}).
	\end{align*}

>>>>

	\begin{proposition}\label{proposition:eigen_noregul}
Let
$
(B^\ast, \Theta, \mathbf{y})
$
denote the observed data defined in \eqref{def:problem_instance}.
At the $l$-th iteration of the PLS algorithm, let
$( \hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top )^\top$
denote the coefficients of $\hat{\xi}$, defined in \eqref{def:maximizer_squared_empirical_cov},
which corresponds to the projected unit-norm direction that maximizes the squared empirical covariance.
These coefficients are obtained by solving the following generalized eigenvalue problem:
\begin{align}
	\left(
	\hat{\mathbf{d}}_1^\top, \ldots, \hat{\mathbf{d}}_K^\top, \hat{\boldsymbol{\zeta}}^\top
	\right)^\top
	=
	\arg \max_{\boldsymbol{\xi} \in \mathbb{R}^{MK+p}}   \boldsymbol{\xi}^\top V \boldsymbol{\xi}
	\quad \text{subject to} \quad \boldsymbol{\xi}^\top B^\ast \boldsymbol{\xi} = 1,
	\label{eq: Regularized generalized rayleigh quotient equation}
\end{align}
where
$$
V := \frac{1}{n^2} (B^\ast \Theta^\top \mathbf{y})(B^\ast \Theta^\top \mathbf{y})^\top \in \mathbb{R}^{(MK+p) \times (MK+p)},
$$

\end{proposition}
Since $V$ is a rank 1 matrix, the solution of the generalized eigenvalue problem is easily obtained by matrix multiplication and normalization:
	\begin{proposition} \label{proposition:linear_noregul}
	We can compute  $\hat{\mathbf{d}}_1, \ldots, \hat{\mathbf{d}}_K$ and $\hat{\boldsymbol{\zeta}}$ in Proposition \ref{proposition:eigen_noregul} as 
\begin{equation*}
	\hat{\mathbf{d}}_j  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	\Theta_j^\top \mathbf{y},
	\quad
	%%%%%%%%%%%%%%%%
	\hat{\boldsymbol{\zeta}}  = \frac{1}{\sqrt{\sum_{j=1}^K 
			\mathbf{y}^\top
			\Theta_j
			B \Theta_j^\top \mathbf{y} 
			+ 
			\mathbf{y}^\top Z Z^\top \mathbf{y}}}
	Z^\top \mathbf{y}.
\end{equation*}
\end{proposition}
The expressions above represent matrix–vector multiplications normalized by a common factor. 
Proofs of Propositions \ref{proposition:eigen_noregul} and \ref{proposition:linear_noregul} are provided in Appendix~\ref{section:proof:proposition:linear_noregul_system_unregularized}.
Although the unnormalized coefficients are computed separately for the functional and scalar part of the PLS component direction, the normalization step couples them, allowing the components to influence each other. This enables the procedure to account for the correlation between functional and scalar components.

	\subsubsection{Proposed Approach: Regularized and Smoothed Computation} \label{sec: Regularization of PLS Components}
	Although Proposition \ref{proposition:eigen_noregul} provides an efficient and principled way to compute the PLS components $\xi$, their functional part $\psi=(\psi^{(1)}, \ldots, \psi^{(K)})$ may not be smooth, which can lead to overfitting and unstable predictions. Thus, in this section, we provide a modification of Proposition  \ref{proposition:eigen_noregul} that yields regularized (smoothed) PLS components.

@@

	Here, $\lambda_k$ $(k=1, \ldots, K)$ is a positive smoothing parameter that controls the trade-off between the goodness of fit and the amount of smoothness in $\psi^{(k)}$. Smaller $\lambda_k$ produces PLS components that fit more closely to the observed data, but selecting too small value may cause overfitting. The estimated PLS component reverts back to the unregularized version of Proposition \ref{proposition:eigen_noregul} when $\lambda_k=0$ for all $k$. On the other hand, larger $\lambda_k$ places more emphasis on smoothing $\psi^{(k)}$. In the limit $\lambda_k \rightarrow \infty$, each of the functions is forced to be of the form $\psi^{(k)}(t)=a+bt$ for some constants $a$ and $b$. In practice, $\{\lambda_k\}_{k=1}^K$ as well as the number of PLS components $L$ to be estimated can be chosen by cross-validation based on a certain predictive performance criterion (e.g., mean squared error).

@@

	\subsection{Proof of Propositions \ref{proposition:eigen_noregul} and \ref{proposition:linear_noregul} }\label{section:proof:proposition:linear_noregul_system_unregularized}

@@

	presented in Proposition \ref{proposition:linear_noregul_system_unregularized} is equivalent to the following constrained optimization problem on the Euclidean space:

@@

	Define the Lagrangian:
	\[
	\mathcal{L}(\{\mathbf{d}_j\}, \boldsymbol{\zeta}, \lambda) 
	:= 
	f(\{\mathbf{d}_j\}, \boldsymbol{\zeta}) - \lambda g(\{\mathbf{d}_j\}, \boldsymbol{\zeta}).
	\]
	The constraint function $g$ is continuously differentiable, and its gradient vanishes only at the origin. Since the origin is not feasible (the constraint requires norm 1), the gradient of $g$ is nonzero at all feasible points. Thus, the linear independence constraint qualification (LICQ) holds, ensuring that the KKT conditions are necessary for local maxima (see, for example, Theorem 5.3.1 in \citealt{bazaraaNonlinearProgrammingTheory2006}).
	Taking gradients of $\mathcal{L}$ and setting them to zero yields:
	\begin{align*}
		\nabla_{\mathbf{d}_j} \mathcal{L} &= 
		\frac{2}{n^2} 
		\left(
		\sum_{\ell=1}^K \mathbf{y}^\top \Theta_\ell B \mathbf{d}_\ell + \mathbf{y}^\top Z \boldsymbol{\zeta}
		\right)
		B \Theta_j^\top \mathbf{y}
		- 2\lambda B \mathbf{d}_j = 0, \quad \forall j, 
		\numberthis \label{lag_grad_d}
		\\
		\nabla_{\boldsymbol{\zeta}} \mathcal{L} &= 
		\frac{2}{n^2} 
		\left(
		\sum_{\ell=1}^K \mathbf{y}^\top \Theta_\ell B \mathbf{d}_\ell + \mathbf{y}^\top Z \boldsymbol{\zeta}
		\right)
		Z^\top \mathbf{y}
		- 2\lambda \boldsymbol{\zeta} = 0.
		\numberthis \label{lag_grad_zeta}
	\end{align*}
	By pre-multiplying equation \eqref{lag_grad_d} by $B^{-1}$ and observing that the scalar quantity $n^{-2}\sum_{\ell=1}^K \mathbf{y}^\top \Theta_\ell B \mathbf{d}_\ell + \mathbf{y}^\top Z \boldsymbol{\zeta}$ appears in both \eqref{lag_grad_d} and \eqref{lag_grad_zeta}, the gradient conditions together imply that any local optimum must be  scalar multiples of
	$
	\Theta_1^\top \mathbf{y}, \ldots, \Theta_K^\top \mathbf{y}, Z^\top \mathbf{y},
	$
	all scaled by the same factor. There exists only one set of vectors  that also satisfies the primal equality constraint, namely the normalized version of the vector set above. The global maximizer is therefore given by
	\begin{equation*}
		\mathbf{d}_j^\ast = \frac{1}{\sqrt{\sum_{j=1}^K 
				\mathbf{y}^\top
				\Theta_j
				B \Theta_j^\top \mathbf{y} 
				+ 
				\mathbf{y}^\top Z Z^\top \mathbf{y}}}
		\Theta_j^\top \mathbf{y},
		\quad
		%%%%%%%%%%%%%%%%
		\boldsymbol{\zeta}^\ast = \frac{1}{\sqrt{\sum_{j=1}^K 
				\mathbf{y}^\top
				\Theta_j
				B \Theta_j^\top \mathbf{y} 
				+ 
				\mathbf{y}^\top Z Z^\top \mathbf{y}}}
		Z^\top \mathbf{y},
	\end{equation*}
	which are the same as  $\hat{\mathbf{d}}_1, \ldots, \hat{\mathbf{d}}_K$ and $\hat{\boldsymbol{\zeta}}$. This completes the proof of  Proposition \ref{proposition:linear_noregul_system_unregularized}.

	\subsection{Proof of Propositions \ref{proposition:eigen_noregul} and \ref{Proposition: Regularized generalized rayleigh quotient} }\label{section:proof:eigenproblem_derivation}
	%
	%
	\begin{align*}
		\widehat{\textnormal{Cov}} (
		\langle \widetilde{\mathbf{W}}, \xi \rangle_{\mathcal{H}}, Y 
		) 
		&= 
		\frac{1}{n} \sum \limits_{i=1}^n 
		y_i
		\langle \widetilde{\mathbf{W}}_i, \xi \rangle_\mathcal{H} \\
		&= 
		\frac{1}{n} 
		\sum \limits_{i=1}^n
		y_i
		\langle \widetilde{X_i}, \psi \rangle_\mathcal{F}  
		+  
		\frac{1}{n} 
		\sum \limits_{i=1}^n  
		y_i
		\langle \mathbf{Z}_i, 
		\boldsymbol{\theta} 
		\rangle 
		\\
		&=  \frac{1}{n} \sum \limits_{i=1}^n  \left\{ \sum \limits_{k=1}^K \int_0^1 X_i(t_k) \psi^{(k)}(t_k) dt_k \right\} Y_i +  \frac{1}{n} \sum \limits_{i=1}^n \mathbf{Z}_i^\top \boldsymbol{\theta} Y_i \\
		&=  \frac{1}{n} \sum \limits_{i=1}^n \left[ \sum \limits_{k=1}^K \int_0^1 \left\{\sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t) \right\} \left\{\sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t) \right\} dt_k \right] Y_i + \frac{1}{n} \sum \limits_{i=1}^n \mathbf{Z}_i^\top \boldsymbol{\theta} Y_i \\
		&= n^{-1} \widetilde{\mathbf{y}}^\top \left[ \sum \limits_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \left\{ \int_0^1 \mathbf{b}^{(k)}(t_k) \mathbf{b}^{(k)}(t_k)^\top dt_k \right\} \mathbf{a}^{(k)} \right] + n^{-1}\widetilde{\mathbf{y}}^\top \mathbf{Z} \boldsymbol{\theta} \\
		&= n^{-1} \widetilde{\mathbf{y}}^\top (\widetilde{C} J \mathbf{a}) + n^{-1}\widetilde{\mathbf{y}}^\top \mathbf{Z} \boldsymbol{\theta} \\
		&= n^{-1} \widetilde{\mathbf{y}}^\top (\widetilde{C}J\mathbf{a} + \mathbf{Z} \boldsymbol{\theta}).
	\end{align*}
